---
title: "OHBM Executive Function Trajectory Demo"
author: "Brenden Tervo-Clemmens"
date: "6/4/2024"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## 1. Data Description and Overview
This is a demonstration of tools for considering statistical and conceptual tests of trajectories across assessments and datasets. The examples used here are from neuropsychological and computerized test of goal-directed cognition or executive function. Theories of human neurobehavioral development suggest executive functions mature from childhood through adolescence reaching adulthood sometime in the second decade of life (e.g., the teenage years in humans). As we (Tervo-Clemmens et al.,2023, Nature Communications) and others have discussed, the development and ultimate maturational timing of such executive functions is often used to demarcate the boundaries of adolescence in clinical practice, research, and policy.

By performing tests across assessments and datasets, we can attempt to draw more robust developmental and lifespan inferences on these questions. 

This demonstration uses fully simulated data (acknowledging privacy and data sharing challenges regarding data use agreements). Variable names and age-related trajectories are based on results from Tervo-Clemmens et al., 2023, Nature Communications. 
Briefly, simulated data used here were generated from simulated uniform age distributions, non-linear trajectories (via GAMs) fit to existing data, and multivariate normal error distributions (approximately matched to observed correlations among executive function measures).

&nbsp;
&nbsp;
&nbsp;

## 2. Load Data and Functions
Here I provide a high level overview and demonstration of some of tools for considering statistical and conceptual tests of trajectories across assessments and datasets. For the sake of streamlining and readability, the more complex code is sourced in a separate file. Let's source these functions and load the first dataset. 
For reproducibility on your machine you will want to change these paths to the relevant location of the cloned git directory. 

## 2.1 Acquire Data
This document and it's supporting materials are hosted on github as [tervoclemmensb/OHBMED_EF](https://github.com/tervoclemmensb/OHBMED_EF). We can download that if it doesn't already exist.

If using these tools or functions please cite Tervo-Clemmens et al., 2023, Nature Communications https://www.nature.com/articles/s41467-023-42540-8

```{r fetchdata, include=TRUE,results='hide'}
if(file.exists("generalfuncs.OHBMdemo.R")) {
   # nothing to do. we're in the directory with R and RData
   cat("already in code repository: ", getwd(), "\n")
} else if(!file.exists("OHBMED_EF/generalfuncs.OHBMdemo.R")) {
   system("git clone https://github.com/tervoclemmensb/OHBMED_EF")
}

# go to the cloned
if(file.exists("OHBMED_EF/generalfuncs.OHBMdemo.R")) setwd("OHBMED_EF") 
```

## 2.2 Load

```{r loaddata, include=TRUE,results='hide', message=F, warning=F}
source("generalfuncs.OHBMdemo.R") # functions: plotfitsfunc
load(file="Simdataset1_EF.Rdata") # "dataset1" object
load(file="Simdataset2_EF.Rdata") # "dataset2" object
```

&nbsp;
&nbsp;
&nbsp;
## 3. Exploring the Data
Let's look at the data by plotting the simulated "raw" data from one test (simulated Design Fluency from the D-KEFS) as a function of age.
```{r explore, nclude=TRUE, message=F, warning=F }
plot(dataset1$simdata1$age,dataset1$simdata1$DFL_acc)
```

We may start to see an existing non-linear trajectory in these data just by visualizing the plot. However, we can of course model this directly. 
Let's model this using the gam function in the mgcv package (https://cran.r-project.org/web/packages/mgcv/index.html) and visualizing the output. Note the example here uses some convenience functions from our sourced functions but we will use a basic smoothed term of age with the default penalized splines via the formula mformula=as.formula('outcome~s(pred)') and we tell the code that age is our predictor variable and DFL_acc is our outcome variable. 
```{r explore2, nclude=TRUE, message=F, warning=F }
##internal call to mgcv is model<-mgcv::gam(mformula,data=dataset1$simdata1)
##convenience function handles model call and prediction (available via sourced functions from above)
DFLaccfit<-mgcvscalefits(dataset1$simdata1,outcomevars = "DFL_acc",predvars = "age",mformula = as.formula('outcome~s(pred)'))
##returned is a dataframe with the fit information
print(DFLaccfit[1:6,1:6])
##plot single fit#####
DFLaccfit$type<-"acc" ##tell convenience function its an accuracy measure for the figure
DFLaccplot<-plotfitsfunc(DFLaccfit,dataset1$variableinfo1,dataset1$simdata1,rawdat=TRUE)
```

&nbsp;
&nbsp;
&nbsp;

Let's look now at the fitted trajectory. 
```{r include=TRUE, message=F, warning=F}
print(DFLaccplot)
```

This plot alone gives us a sense of the potential developmental inferences at the group level. For example, we see support for neurodevelopemntal theories of performance developmentally stabilizing towards the end of the second decade of life.  

Of course, our current goal is to consider broader inferences across multiple measures. 

&nbsp;
&nbsp;
&nbsp;

## 4. Model Fits from Multiple Assessments

We can visualize multiple developmental processes by repeating the processes from above and stacking the derived model fits. Again, by performing tests across assessments of the same broad construct of executive function, we can draw more robust developmental and lifespan inferences. 

```{r include=TRUE, message=F, warning=F}
###get all the outcomes 
alldataset1outcomes<-names(dataset1$simdata1)[names(dataset1$simdata1)!="age"]
print(alldataset1outcomes)

fitsalldataset1<-mgcvscalefits(dataset1$simdata1,outcomevars = alldataset1outcomes,predvars = "age",mformula = as.formula('outcome~s(pred)'))

```

Let's now visualize all the fits. We will remove the individual data points for clarity and provide a shape to the lines so we know which model fit goes with which measure. 

```{r include=TRUE, message=F, warning=F}
dataset1plots<-plotfitsfunc(fitsalldataset1,dataset1$variableinfo1,dataset1$simdata1,points=TRUE)
print(dataset1plots)
```

We can visualize the consistency across measures in the age trajectory. From a lifespan perspective, this helps understand similarities in the developmental process. We can also see consistency towards the previously discussed developmental theories on rapid development during childhood and early adolescence that stabilizes during the second decade of life. However, we may also be interested in interrogating specifics about the developmental timing and shape of these trajectories. 

&nbsp;
&nbsp;
&nbsp;

## 5. Developmental Inference from Derivatives

Ideally we would make quantitative inferences about developmental timing and the shape of trajectories. One way that we and others (e.g., Simmonds et al., 2014, NeuroImage; Calabro et al., 2020, Cerebral Cortex; Tervo-Clemmens et al., 2023, Nature Communications) have done this is to statistically examine the first derivative of the fitted trajectories. 

Let's return to our first example of design fluency. For simplicity we will remove the raw data points and just focus on the trajectory. 

```{r include=TRUE, message=F, warning=F}
DFLaccplot2<-plotfitsfunc(DFLaccfit,dataset1$variableinfo1,dataset1$simdata1,rawdat=FALSE)
print(DFLaccplot2)
```

Using tools from the gratia package (https://cran.r-project.org/web/packages/gratia/index.html) we can also examine the first derivative of this trajectory and its confidence interval (via posterior simulation: see package for more details). See the sourced functions for more details beyond these convenience functions. 

Here we plot the first derivative and its 95% CI. This plotting function will provide color (red if positive, blue if negative) for periods where the CI does not include zero. 

```{r include=TRUE, message=F, warning=F}
DFLderivdat<-mgcvgam_growthrate_multiplot(df=dataset1$simdata1,outcomevars = "DFL_acc",predvars='age',mformula=as.formula('outcome~s(pred)'))
DFLderivdatplot<-derivplot_unthresholded(DFLderivdat$derivdata)
print(DFLderivdatplot)
```

&nbsp;
&nbsp;
&nbsp;

&nbsp;
&nbsp;
&nbsp;

Instead of looking at the full derivative and its CI, we might also make this a 1D heat plot, where just utilize the color, where again color is displayed when its significant and no color displayed when its not. We and others have used such maps of significance of derivative to chart lifespan processes. For example, we can look at the ages when there is significant change and when this stops as an insight into developmental processes. We will provide a second bar (all white) below to anchor this developmentally for key periods at 10, 12, 15, 18 years old. 

&nbsp;
&nbsp;

```{r include=TRUE, message=F, warning=F}
DFLderivdatplotraster<-mgcvgam_growthrate_multiplot(df=dataset1$simdata1,outcomevars = "DFL_acc",predvars='age',mformula=as.formula('outcome~s(pred)'),devageguides=TRUE)
print(DFLderivdatplotraster$returnplot)

```

This further gives us insight and support for developmental theory wherein significant age-related changes occur during early in adolescence that then stabilize (i.e., no more age-related change) around 18-years-old.

&nbsp;
&nbsp;
&nbsp;

## 5. Derivatives of Multiple Assessments

In addition to looking at the derivative for a single executive function measure, here our emphasis is on drawing conclusions across mutiple measures. Let's repeat the process now for all of the measures. We will plot the accuracy and latency measures separate for consistency. 

&nbsp;
&nbsp;
Accuracy Measures
```{r include=TRUE, message=F, warning=F}
###dataset1
dataset1accvars<-dataset1$variableinfo1$outcome[dataset1$variableinfo1$type=="acc" & !is.na(dataset1$variableinfo1$outcome)]
ggderivplotsacc<-mgcvgam_growthrate_multiplot(df=dataset1$simdata1,outcomevars = dataset1accvars,predvars='age',mformula=as.formula('outcome~s(pred)'))
print(ggderivplotsacc$pairswithfill)
print(ggderivplotsacc$returnplot)
```
&nbsp;
&nbsp;
Latency Measures
```{r include=TRUE, message=F, warning=F}
dataset1latvars<-dataset1$variableinfo1$outcome[dataset1$variableinfo1$type=="lat" & !is.na(dataset1$variableinfo1$outcome)]
ggderivplotslat<-mgcvgam_growthrate_multiplot(df=dataset1$simdata1,outcomevars = dataset1latvars,predvars='age',mformula=as.formula('outcome~s(pred)'))
print(ggderivplotslat$pairswithfill)
print(ggderivplotslat$returnplot)
```

Supporting the similarities of the fits, the derivatives make clear the consistency of the maturational timing (i.e., when no more development is achieved) of this distinct measures. However, this is a single dataset, what about integrating information and inferences across datasets. 

&nbsp;
&nbsp;
&nbsp;

## 5. Chracterizing Trajectories across Datasets

Let's load another dataset. Again, this is a simulated data (now from a second independent dataset), with variable names and age trajectories based on our previous work. See section 1 for more information. 

We will load the data, fit the models for all the measures, and visualize by repeating steps from above. 
```{r include=TRUE, message=F, warning=F}
#Dataset 2
alldataset2outcomes<-names(dataset2$simdata2)[names(dataset2$simdata2)!="age"]
print(alldataset2outcomes)

fitsalldataset2<-mgcvscalefits(dataset2$simdata2,outcomevars = alldataset2outcomes,predvars = "age",mformula = as.formula('outcome~s(pred)'))

dataset2plots<-plotfitsfunc(fitsalldataset2,dataset2$variableinfo2,dataset2$simdata2)
print((dataset2plots+ggtitle("Dataset 2 Fits")+theme(plot.title = element_text(hjust = 0.5))))
```

&nbsp;

Bringing back the fits from dataset 1, we can the high degree of similarity of between the two datasets, which gives us more confidence in our overall inferences regarding our developmental inferences. 

```{r include=TRUE, message=F, warning=F}
print((dataset1plots+ggtitle("Dataset 1 Fits")+theme(plot.title = element_text(hjust = 0.5))))
```

&nbsp;

We likewise see the same general pattern with the derivatives. For example, here are the accuracy trajectories from both datasets. 

&nbsp;

```{r include=TRUE, message=F, warning=F}
dataset2accvars<-dataset2$variableinfo2$outcome[dataset2$variableinfo2$type=="acc" & !is.na(dataset2$variableinfo2$outcome)]
dataset2latvars<-dataset2$variableinfo2$outcome[dataset2$variableinfo2$type=="lat" & !is.na(dataset2$variableinfo2$outcome)]

ggderivplotsacc2<-mgcvgam_growthrate_multiplot(df=dataset2$simdata2,outcomevars = dataset2accvars,predvars='age',mformula=as.formula('outcome~s(pred)'))
print(ggderivplotsacc2$returnplot+ggtitle("Dataset 2 Accuracy Measures Derivatives")+theme(plot.title = element_text(hjust = 0.5)))
```

```{r include=TRUE, message=F, warning=F}
print(ggderivplotsacc$returnplot+ggtitle("Dataset 1 Accuracy Measures Derivatives")+theme(plot.title = element_text(hjust = 0.5)))

```

These inferences are helpful from a general reproducibility perspective to support the proposed developmental theories. In the broader presentation we will put these into more context and preview further potential examples and extensions of these foundational approaches.
